Reconstructing the last known movements of one of Nature's giants

### 2017-11-15 manuscript/revision/response.md

The abstract was clear and direct, providing a good summary of the paper's approach and conclusions.
The conclusion is clear and justified, although the claim of first quantitative evidence of movement behaviours in blue whale seems unlikely given the slightly older paper from Busquets-Vass also reporting blue whale baleen isotopes.

> We have removed this comment from the manuscript, and discuss the novelty of our work more in the response to referee 2 below.

## Referee 2:

> If we are interpreting Referee 2's comments correctly (see below), we think they mostly revolve around issues about the novelty of the study, and overstatement of results based on one individual. We completely see the points the referee is making. Some of this was related to the difficulties of writing in the shorthand Nature style - we have clearly been a little too casual in some of our wording. In other cases we haven't been clear that the selling point of this method is that it can easily be applied to hundreds of individuals at relatively low cost. Here we test the methods on one individual, but the real value in the method is its broader applicability. We have tried to clarify this where possible and have also toned down/removed/clarified a lot of our previous statements about the novelty of the study and interpretations of the results. We explain our changes in more detail below, and apologise for any offense caused by our apparent disregard for previous excellent work in this area.

@@

The paper is interesting and as the authors state, provides potentially valuable information on (one) historical blue whale movements in the North Atlantic Ocean. However, I have substantial concerns regarding the originality/novelty, assumptions, and conclusions of the study, enough so that I cannot recommend its publication in Nature.

The Abstract and text explicitly or implicitly state the novelty of the following: “first quantitative multi-year records of historical migrations in individual blue whales”, similarity of blue whale movements over hundreds of years, use of archived museum specimens, novel agent based simulation models.

The study does not provide "multi-year records of...individual blue whales", but of one blue whale. The generalization of the results from one individual to the entire population is a central issue to the current paper.

> This is a fair point, and partially the result of some lax wording on our part. We have toned down our claims of novelty throughout and removed this sentence from the abstract.

@@

There are also issues with the movement interpretations for this one whale; the most important is that the authors make strong conclusions re: movement when the measured SIA values match those predicted by their model, but then employ a physiological argument for 'phase 2' (pregnancy and lactation) when there is a mismatch between measured data and the model. Of course, this approach makes the assumption that physiology can override the environmental signal, which calls into question any inferred movement patterns throughout the time series.

> The referee is correct in that

A main conclusion of the paper, that blue whales then migrated as they do now, simply cannot be made from one whale. I appreciate the difficulty of obtaining specimens like the one here, which makes larger sample sizes difficult to obtain. That limitation does make n = 1 studies reasonable, worthwhile, and interesting; it is my opinion that the results here, both for insights into the movements of this whale and for the methodology employed (combining a movement/isoscape model with measured values), should be published. However, n = 1 studies cannot be used to generalize the movements of an entire wild, historical population.

> This is a fair point again. As stated above this is partially an issue of our wording and attempts to write this in a Nature style. We have toned down these comments throughout, and focus instead on promoting the methodological approach we have used, which we do believe is novel.

@@

This paper is also not the first to use historical data to reconstruct past movements of blue whales; see for example Reeves et al (2004) that addresses the same question with whale landings data.

As the authors cite, the reconstruction of whale movements using baleen is well established in the literature; see papers by Best, Schell, Hobson, and others for similar analyses on Bowhead, Right, and other baleen whales.

> Yes we are aware of these other papers, and cite many of them already. We agree that our wording in the original manuscript was not very clear and overly stated the novelty of our findings. We have tried to tone this down throughout, and provide more balanced coverage of previous work.

@@

The penultimate sentence of the abstract and final sentence of the paper suggest that using archived museum specimens is a novel concept in marine ecology. The use of archived museum specimens is not new or overlooked; two immediate examples (of many) include Vo et al (2010) using century-old seabirds and Yeakel et al (2009) on the famous lions of Tsavo. There are papers written that point to the value of using archived collections for data (e.g. Lister 2011), and the approach is common enough that there is a methods paper written on the preparation of archived specimens for SIA (Edwards et al 2002).

> Again we realise that our wording was lax in places and overstated the novelty of this approach. This was not our intention - our intention was to encourage more ecologists to consider these archives of data in their work. We are often surprised by how under-utilised our collections are for this kind of work (even in the NHM) so wanted to highlight this. We have toned down our claims of novelty and have added more references to clarify that this is an established field of research.

@@

The “novel” movement model using SIA is indeed an improvement over many past studies using SIA to reconstruct animal movements. However, there are some studies have employed similar methods (e.g. Flockhart et al 2013 in Monarch butterflies).

> OK ???? HAVE THEY???

In short, the authors applied a common tool in an established way to a single historical specimen of a rare species (though their agent-based movement model is a significant improvement over similar studies). Their model matched their measurements sometimes, but not always, leading to a movement-based interpretation for one period and a physiology-based interpretation (pregnancy and lactation) for the second period.

Perhaps most importantly, if the results and conclusions are considered completely correct, the authors still provide the movements of only a single whale. The conclusion that historical blue whale movements were the same as today's based on a single individual is simply not appropriate.

> We have removed this from the text and have toned down interpretations that relate to just a single individual.

@@

> We hope our edits have addressed these issues.

See below for additional, specific comments.

Abstract: the inference of pregnancy and lactation seems more questionable than as it is stated here.

> We have removed the reference to lactation, and reworded to make it clear that the pregnancy conclusion is more tenuous than it appeared in the original abstract.

@@

As mentioned above, the Abstract overextends the results (e.g., "migrations in blue whales" shouldnt be plural; "hundreds of years" which is not the timeline they present, etc).

> We have reworded the abstract and removed this inaccurate comment.

@@

Intro: migratory species may be affected by climate change by modifying movements, but aren't necessarily 'vulnerable'. Some would argue that migratory species are the least vulnerable to climate change.

> We have reworded this to "Such species may also be particularly affected by climate change".

@@

lines 51-52 criticize approaches that are limited to only a few individuals, but this paper provides information on only one individual.

> Fair point. We were trying to clarify that our approach is more scalable, but agree that this was poorly worded. Instead we now state on line XXX: "These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum (NHM) London's blue whale `Hope'."

@@

Fig 1 is not very clear. There is no clear indication of the whales movement, and no isotope data on the Figure. I understand the isoscape is in the SM but this figure could be improved to provide better movement and/or isotope data.

> We are a little confused by this comment. I assume the referee meant Figure 2? Figure 1 is a trace of d13C and d15N, and quite clearly contains isotope data. 
FIX!!!

@@

The movement simulations in Fig 2 are also confusing. The SM clarifies but this Figure could show more. For example, the data matches the simulations...of what? The reader cannot tell if the differences in simulations are due to isotopic variability or different movement scenarios. Perhaps the 4 movement scenarios shown in SM could be 4 different colors here. In that case, when the measured data match a given slope, peak, or trough, the reader can associate it with a geographic region.

> Again, we assume the referee is referring to Fig 3 not Fig 2. This figure has changed now we have changed our movement models and added more data. We have clarified what the figure is showing in the legend as follows: :
FIX!!!

@@

lines 112-122. The conclusion of an isotopic reflection of pregnancy and lactation is stated rather strongly here given the evidence (the authors describe the inference as 'tentative' in the SM).

> We have rephrased this paragraph in response to the new data and analyses we have added. We have also toned down the claims of pregnancy and removed those about lactation.

@@

128: "resident" to where?

>

137: this point made extensively above....describes the overall "spatial ecology of blue whales in the North Atlantic" based on isotopic inferences over 6 years in a single whale.

> We have removed this statement and toned down claims based on this one whale throughout.

### 2017-11-15 manuscript/revision/blue-whale-ms-v2.tex
Edits to main text in response to referee 2

\raggedright
% Need to edit
\paragraph{Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}. 
Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London. 
We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
Subtropical residency is associated with isotopic signatures consistent with pregnancy in the year before the whale's death. 
These results supply the first quantitative multi-year records of historical migrations in individual blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region. 
Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history. 
Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.}

%\section{Main text}
\newpage

Migratory species pose a particular challenge for conservation practitioners, because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
Such species are also likely to be particularly vulnerable to climate change, as changes may impact various parts of their life cycle\cite{robinson2009travelling}. 
Dealing with these issues is difficult due to the relative scarcity of information on individual-level movements over multiple years, for both historical and present-day populations, especially for wide-ranging marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.

The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List\cite{reilly2008balaenoptera}.
Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, both for historical and present-day populations. 
Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}. 
These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}. 
Stable isotope analyses provide an alternative technique for investigating blue whale movements\cite{west2006stable}. 
We can use the isotopic compositions of whale tissues to reconstruct their movement patterns, by comparing them to spatially referenced models of isotopic variation across the species' range\cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}. 
%line 57,
These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale `Hope'.

%Points here about movement patterns, interactions with human use of oceans, tourism etc
% Build on lack of information about movements before ,modern use of Atlantic, and lack of information about juvenile behaviour.
% Introduce historic samples as archive behavoiiral information

Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
It is ideal for stable isotope studies because keratin grows continuously through an individual's life (although it is worn away at the tips over time, so rarely records an individual's entire lifespan), and once laid down it is metabolically inert\cite{best1996stable}. 
Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around two weeks prior to that time (two weeks is required for the carbon and nitrogen to be assimilated and converted into tissues). %REF???
Baleen stable isotopes have a long history of use as biochemical markers to infer movement patterns in whales (REFS). % ADD REFS
%Refs to support long history of inferring movement from baleen isotopes.

We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth. 
This resulted in XX discrete samples of baleen. % add number
We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin. 
These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e. relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope\cite{west2006stable}.
Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; isotopes provide information on the geographical locations of an individual, whereas nitrogen isotopes provide trophic and physiological information\cite{west2006stable}.

Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the seven full years of the whale's life (early 1884 - spring 1891). 
The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods).  
In behavioural phase one (from record start in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig. \ref{fig1}; Figs. S1 and S2). 
In behavioural phase two (summer 1886 to spring 1891) $\delta^{15}$N values are relatively elevated and we see coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values. 
In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive $\delta^{13}$C values in the second half of 1891. 
The final three months of the record show a continued fall in $\delta^{13}$C values (Fig. \ref{fig1}; Figs. S1 and S2).

@@

% figure 1
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis). Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes). There is strong annual periodicity and cross-correlation (Figs. S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1889 (approximately at sample 30cm). Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  \label{fig1}
\end{figure}

@@

Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time \cite{west2006stable,mcmahon2015millennial}.
We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig. S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig. S5) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales. 
We then compare simulated profiles (Figs. \ref{fig2}, \ref{fig3} and Fig. S4) with the measured isotopic profile of the blue whale's baleen (Fig. \ref{fig1}) to identify the most likely movement history.
We assume continued ingestion of new protein throughout the year. 
Other mysticete whales fast during migrations and during winter, relying on stored lipid reserves to meet metabolic demands, but blue whales are too large to stop feeding for long periods; \cite{busquets2017estimating,silva2013north}).
%Can we back this up? Are we more saying that we think they need protein to maintain tissue turnover but could meet energetic demands form stored lipids?

@@

% figure 2
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-2-migratory-model-full-map.png}
  \caption{Simulated tracks from the migratory movement model (2250 days for each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Norwegian/Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow). }
  \label{fig2}
\end{figure}

% figure 3
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-3-migratory-model-d13C.png}
  \caption{Correlations among $\delta^{13}$C data from 50 movement simulations (green) and empirical data from baleen (blue). Across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 ($sd = 0.2$), 49 of them were positive, and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected value of 0.001 ($p = 0.05/50 = 0.001$). The initiation of the southward migration in the late spring of 1889 is indicated by the vertical broken line. The standard deviation of the empirical data have been rescaled by a factor of three so as to better visualise the correlation because, as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
  \label{fig3}
\end{figure}

@@

We simulated 1000 individual movement patterns and compared resulting simulated baleen $\delta^{13}$C records to the measured records with simple linear regressions. 
Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1000 simulated profiles was 0.49, and the maximum was 0.74 (Fig. S6, distribution of $r^2$ values). 
The top 10\% best fitting simulated profiles are shown in Fig. \ref{fig2} [line plot top 10\%). 
Best fitting models predict juvenile residency in the Cape Verde region in behavioural phase 1. 
Behavioural phase 2 is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and sub-tropical waters. 
Best fitting models in general predicted a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary materials Fig. S7 [boxplots and kernel density plots]).

%line 112 - 122 para
The last 500 days of the whale's life are difficult to simulate. 
Beginning in the winter of 1889/1890, the observed $\delta^{13}$C profile is relatively depleted, and remains constant for approximately six months, before increasing rapidly in the winter of 1890/1891. 
Depleted $\delta^{13}$C values are found in northern waters that show large temporal fluctuations in $\delta^{13}$Cplk values [fig SX residency plot], and the observed values cannot be simulated purely from movements within the known geographic range.
We infer that the $\delta^{13}$C record in the last period of life reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}. 
We suggest that the constant, depleted $\delta^{13}$C values reflect use of carbon reserves assimilated in northern latitudes during 6-8 months residency in low latitude waters. 
After six months the whale began feeding in low latitude waters producing a dramatic increase in $\delta^{13}$C values and began a final migration north. 
Blue whales have a 10-12 month gestation period, with calving occurring in sub-tropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, residency in warm waters through the whole of 1890 with feeding resuming in the second half of 1890 and eventual stranding during the return to northern feeding grounds in early 1891, is therefore consistent with the known reproductive ecology of blue whales.

>>>>

\raggedright
% Need to edit

\section{Abstract}
Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}. 
Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London. 
%We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
%Subtropical residency is associated with isotopic signatures consistent with pregnancy in the year before the whale's death. % make clear this is tenuous.
% Juvenile behaviour?

Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history. 
Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.

\newpage
\section{Main text}
Migratory species pose a particular challenge for conservation practitioners, because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
Such species may also be particularly affected by climate change, as changes can impact various parts of their life cycle\cite{robinson2009travelling}. 
Dealing with these issues is difficult due to the relative scarcity of information on individual-level movements over multiple years, for both historical and present-day populations, especially for wide-ranging marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.

The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List\cite{reilly2008balaenoptera}.
Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, particularly for juveniles. 
Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}. 
These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}. 
They also cannot provide information on historical movement patterns of blue whales prior to overexploitation and modern use of the oceans, which could be important for understanding the drivers of their declines.
An alternative technique for investigating blue whale movements in both present-day and historical populations is to use stable isotopes taken from individuals in the field or from museum specimens\cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
Historic samples from museum collections form a rich archive of behavioural information \cite{lister2011natural} that is often under-utilised.
We can use the isotopic compositions of whale tissues to reconstruct their movement patterns, by comparing them to spatially referenced models of isotopic variation across the species' range\cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}. 
These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale `Hope'.

Baleen stable isotopes have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}.
Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
It is ideal for stable isotope studies because keratin grows continuously through an individual's life (although it is worn away at the tips over time, so rarely records an individual's entire lifespan), and once laid down it is metabolically inert\cite{best1996stable}. 
Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around two weeks prior to that time (two weeks is required for the carbon and nitrogen to be assimilated and converted into tissues). %REF???

We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth. 
This resulted in XX discrete samples of baleen. % add number
We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin. ,
These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e. relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope\cite{west2006stable}.
Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; carbon isotopes provide information on the geographical locations of an individual, whereas nitrogen isotopes provide trophic and physiological information\cite{west2006stable}.

Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the seven full years of the whale's life (early 1884 - spring 1891). 
The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods).  
In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig. \ref{fig1}; Figs. S1 and S2). 
In behavioural phase two (summer 1886 to spring 1891) $\delta^{15}$N values are relatively elevated and we see coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values. 
In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive $\delta^{13}$C values in the second half of 1891. 
The final three months of the record show a continued fall in $\delta^{13}$C values (Fig. \ref{fig1}; Figs. S1 and S2).

@@

% figure 1
%\begin{figure}
 % \centering
  %\includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
  %\caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis). Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes). There is strong annual periodicity and cross-correlation (Figs. S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1889 (approximately at sample 30cm). Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  %\label{fig1}
%\end{figure}

@@

Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time \cite{west2006stable,mcmahon2015millennial}.
We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig. S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig. S5) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales. 
We then compare simulated profiles (Figs. \ref{fig2}, \ref{fig3} and Fig. S4) with the measured isotopic profile of the blue whale's baleen (Fig. \ref{fig1}) to identify the most likely movement history.
We assume continued ingestion of new protein throughout the year. 
Other mysticete whales fast during migrations and during winter, relying on stored lipid reserves to meet metabolic demands, but blue whales are too large to stop feeding for long periods; \cite{busquets2017estimating,silva2013north}.
%Can we back this up? Are we more saying that we think they need protein to maintain tissue turnover but could meet energetic demands form stored lipids?

@@

% figure 2
%\begin{figure}
 % \centering
 % \includegraphics[width = \linewidth]{figures/Figure-2-migratory-model-full-map.png}
  %\caption{Simulated tracks from the migratory movement model (2250 days for each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Norwegian/Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow). }
  %\label{fig2}
%\end{figure}

% figure 3
%\begin{figure}
 % \centering
  %\includegraphics[width = \linewidth]{figures/Figure-3-migratory-model-d13C.png}
  %\caption{Correlations among $\delta^{13}$C data from 50 movement simulations (green) and empirical data from baleen (blue). Across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 ($sd = 0.2$), 49 of them were positive, and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected value of 0.001 ($p = 0.05/50 = 0.001$). The initiation of the southward migration in the late spring of 1889 is indicated by the vertical broken line. The standard deviation of the empirical data have been rescaled by a factor of three so as to better visualise the correlation because, as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
  %\label{fig3}
%\end{figure}

@@

We simulated 1000 individual movement patterns and compared resulting simulated baleen $\delta^{13}$C records to the measured records with simple linear regressions. 
Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1000 simulated profiles was 0.49, and the maximum was 0.74 (Fig. S6, distribution of $r^2$ values). 
The top 10\% best fitting simulated profiles are shown in Fig. \ref{fig2} [line plot top 10\%). 
Best fitting models predict juvenile residency in the Cape Verde region in behavioural phase 1. 
Behavioural phase 2 is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and subtropical waters. 
Best fitting models in general predicted a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary materials Fig. S7 [boxplots and kernel density plots]).

%line 112 - 122 para
The last 500 days of the whale's life are difficult to simulate. 
Beginning in the winter of 1889/1890, the observed $\delta^{13}$C profile is relatively depleted, and remains constant for approximately six months, before increasing rapidly in the winter of 1890/1891. 
Depleted $\delta^{13}$C values are found in northern waters that show large temporal fluctuations in $\delta^{13}$Cplk values [fig SX residency plot], and the observed values cannot be simulated purely from movements within the known geographic range.
We infer that the $\delta^{13}$C record in the last period of life reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}. 
We suggest that the constant, depleted $\delta^{13}$C values reflect use of carbon reserves assimilated in northern latitudes during 6-8 months residency in low latitude waters. 
After six months the whale began feeding in low latitude waters producing a dramatic increase in $\delta^{13}$C values and began a final migration north. 
Blue whales have a 10-12 month gestation period, with calving occurring in subtropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, residency in warm waters through the whole of 1890 with feeding resuming in the second half of 1890 and eventual stranding during the return to northern feeding grounds in early 1891, is therefore consistent with the known reproductive ecology of blue whales.

%%% STILL TO FIX BELOW %%%%
